\documentclass{amsart}

\input{C:/Users/jzchr/MathDocuments/preamble_general.tex}

\title{Math 325 HW 2}
\author{Jalen Chrysos}

\begin{document}
	
\noindent\textbf{Problem 1}: Let $G$ be a subgroup of $\GL_n(k)$ and 
	$$
	I := \bigoplus_{d>0} P_d^G, \;\;\;\; I_D := \bigoplus_{d>0} (D_d(\R^n))^G
	$$
	(note that we exclude constants). A polynomial $f\in \C[x_1,\dots,x_n]$ is called $G$-\textit{harmonic} if $u(f)=0$ for all $u\in I_D$.
	\begin{enumerate}[(a)]
		\item Show that $f$ is $G$-harmonic iff $\<u_1u_2,f\>=0$ for all $u_1\in D(\R^n)$ and $u_2\in I_D$.
		\item (optional) Let $H$ be the vector space of $G$-harmonic polynomials. Prove that if $G$ is a subgroup of $\OO_n$, then 
		$$
		\C[x_1,\dots,x_n] = H\oplus \C[x_1,\dots,x_n]I.
		$$
		\item Show that $f$ is $\SO_n$-harmonic iff $\Delta(f)=0$ and interpret the statement of (ii) in this case.
	\end{enumerate}
	
	\begin{proof}
		(a): If $\<u_1u_2,f\>=0$ for all $u_1\in D(\R^n)$ then in particular for $u_1=1$, $\<u_2,f\>=0$.
		
		Conversely, suppose $u(f)=0$ for $u\in I_D$. $f$ can be split into $f_0+f_1+\dots+f_d$ where $f_k\in P_k$. \textit{to do}\\
		
		(c):  
	\end{proof}
	
\newpage 

\noindent\textbf{Problem 2}: Let $A$ be a $k$-algebra and $a$ be algebraic over $A$ with minimal polynomial $p_a$. Show that $\Spec(a)$ is the set of roots of $p_a$ in $k$. 

\begin{proof}
	$\Spec(a)$ is defined as the set of $\lambda \in k$ for which $(a-\lambda)$ is invertible. We use the fact that 
	$$
	p_a(t) - p_a(\lambda) = (t-\lambda)q(t)
	$$
	for some polynomial $q(t)$ with degree $\deg(p_a)-1$. In the case $t=a$, this gives
	\begin{align*}
	p_a(a) - p_a(\lambda) &= (a-\lambda)q(a) \\
	 -p_a(\lambda) &= (a-\lambda)q(a)\\
	 (a-\lambda)^{-1} &= -p_a(\lambda)^{-1}q(a)
	\end{align*}
	(note that $q(a)\neq 0$ since $p_a$ is minimal and $\deg(q)<\deg(p_a)$). So as long as $p_a(\lambda)\neq 0$, $\lambda \not\in \Spec(a)$. However, if $p_a(\lambda)=0$, then $(a-\lambda)$ is not invertible because it divides $p_a(\lambda)$ i.e. it is a zero-divisor.
\end{proof}

\newpage
\noindent \textbf{Problem 3}: Let $A$ be a $k$-algebra and $M$ a \textit{simple} $A$-module. Prove that 
$$\dim_k(\End_A M) \leq \dim_k M \leq \dim_k A.$$
(here $\leq$ is the order on cardinals).

\begin{proof}
	Pick some nonzero $m\in M$ and let $f:\End_A M \to M$ be defined by $f:\ph\mapsto \ph(m)$. $f$ is injective as follows: suppose $\ph,\psi\in \End_A M$ and $\ph(m)=\psi(m)$. Then define $M'\subseteq M$ by
	$$
	M' := \{m'\in M: \ph(m')=\psi(m')\}.
	$$
	Because $\ph$ and $\psi$ are linear, $M'$ is a submodule of $M$. But $M$ is simple, so $M'$ is either $0$ or $M$. If it is $0$ then no such $\ph,\psi$ exist, thus $f$ is injective. If $M'=M$, then $\End_A M$ is trivial. In either case, $\dim_k(\End_AM) \leq \dim_k M$.\\
	
	For the other inequality, let $g:A\to M$ be the map $g:a\mapsto a\cdot m$. The image of $g$ is a submodule of $M$, but $M$ is simple so $\im(g)$ must be $0$ or $M$ and it is clearly not $0$ because $g(1)=m\neq 0$. Thus $g$ is surjective, establishing that $\dim_kM \leq \dim_k A$.
\end{proof}\\

\newpage
\noindent \textbf{Problem 4}: Let $A$ be a ring with nonzero idempotent $e=e^2$. Note that the set $eAe$ forms a ring with unit $e$. Prove the following:
\begin{enumerate}[(a)]
	\item For any left $A$-module $M$, there is a natural $eAe$-module structure on $eM$. Furthermore, for any $A$-modules $M,N$, there is a natural morphism of additive groups $f:\Hom_A(M,N)\to \Hom_{eAe}(eM,eN)$.
	\item Multiplication on the right gives an algebra isomorphism $(eAe)^{\rm{op}} \to \Hom_A(Ae,Ae)$.
	\item Assuming that $AeA=A$, the map $f$ in (a) is a bijection.
	\item Assuming $AeA=A$, an $A$-module $M$ is simple iff the $eAe$-module $eM$ is simple (and similarly for direct sums of simple modules).
	\item (optional) Assuming $AeA=A$, there is an algebra isomorphism $A^{\rm{op}}\to \Hom_{eAe}(eA,eA)$.
\end{enumerate}

\begin{proof}
	(a): To give a module structure we have to specify how scaling and addition work. They are as follows:
	$$
	eae \cdot em = e(aem) \in eM, \;\;\; em_1 + em_2 = e(m_1+m_2)\in eM.
	$$
	For the morphism $f:\Hom_A(M,N)\to \Hom_{eAe}(eM,eN)$, take
	$$
	f: \ph \mapsto \big(em \mapsto e \ph(m) \in eN\big).
	$$
	To verify that $f(\ph)$ is indeed $eAe$-linear, 
	$$
	f(\ph)(eae \cdot em) = e \ph(aem) = eae \ph(m) = eae\cdot  f(\ph)(em)
	$$
	$$
	f(\ph)(em_1 + em_2) = e\ph(m_1+m_2) = e\ph(m_1) + e\ph(m_2)=f(\ph)(em_1) + f(\ph)(em_2).
	$$\\
	
	(b): For $ebe\in eAe$, we have the associated linear map $Ae\to Ae$ given by
	$$
	\Phi(ebe) = ae \mapsto ae \cdot ebe = (aeb)e \in Ae.
	$$
	This map $\Phi:(eAe)^{\rm{op}}\to \Hom_A(Ae,Ae)$ is an algebra homomorphism; it is clearly linear, and to show that it is multiplicative,
	$$
	\Phi(ebe) \circ \Phi(ece) = \big(ae \mapsto ae\cdot ece \cdot ebe = aecebe\big) = \Phi(ece \cdot ebe)
	$$
	(note that the order is switched, reflecting the oppositeness of $(eAe)^{\rm{op}}$).
	To show that it is an isomorphism, we have the inverse 
	$$
	\Phi^{-1}: \ph \mapsto e\ph(e)
	$$ 
	and one can check that $\Phi,\Phi^{-1}$ are indeed inverses:
	$$
	(\Phi^{-1}\circ \Phi)(ebe) = \Phi^{-1}\big(ae \mapsto (aeb)e\big) = eeebe = ebe
	$$
	and 
	$$
	(\Phi\circ \Phi^{-1})(\ph) = \Phi(e\ph(e)) = \big(ae \mapsto ae e\ph(e) = \ph(ae)\big) = \ph.
	$$\\
	
	(c): With the assumption that $AeA=A$, there is some linear combination $\sum_{i=1}^n a_ieb_i = 1$ where $a_i,b_i\in A$. Thus, any $m$ can be expressed as a $A$-linear combination of elements in $eM$:
	$$
	m = \sum_{i=1}^n a_i\cdot e(b_im).
	$$
	Using this idea, we can define the inverse of $f$ as
	$$
	f^{-1}: \ph \mapsto \big(m \mapsto \sum_{i=1}^na_i \cdot \ph(eb_im)\big).
	$$
	It is simple to check that these are inverses.\\
	
	(d): By (c), homomorphisms $M\to N$ are in bijection with those $eM\to eN$, thus $M$ is a nontrivial submodule of $N$ iff $eM$ is of $eN$, and hence $M$ is simple iff $N$ is simple.\\
	
	
	
\end{proof}
	
\end{document}
