\documentclass{amsart}
\input{C:/Users/jzchr/MathDocuments/preamble_general.tex}

\title{Differentiable Manifolds}
\author{Jalen Chrysos}

\begin{document}
	
\maketitle

These are my notes for Topology II taught by Eduard Looijenga at UChicago in Winter 2026.

\tableofcontents

\newpage 

\section{Basic Definitions}

A \textit{topological manifold} of dimension $m$ is a topological space $M$ that is Hausdorff and locally homeomorphic to $\R^m$. Such an $M$ has an open covering $\A = \{U_{\a}\}$ called an \textit{atlas} with associated homeomorphisms (\textit{charts}) $\kappa_{\a}:U_{\a}\to \R^m$ which are compatible, meaning that in each intersection $U_{\a}\cap U_{\b}$, we have a homeomorphic coordinate change map:
$$
\R^m \supset \kappa_{\a}(U_{\a}\cap U_{\b}) \xra{\kappa_{\b}\kappa_{\a}^{-1}} \kappa_{\b}(U_{\a}\cap U_{\b}) \subset \R^m
$$
The atlas is $C^k$ if all the coordinate change maps are $C^k$.\\

$\A$ must be $C^k$ in order to define the notion of a $C^k$ function $M\to \R$ (relative to $\A$); otherwise, we could have $f:M\to \R$ that is $C^k$ through one chart but not another. Naturally, which functions $M\to \R$ are $C^k$ depends on $\A$. And in fact, \textbf{atlases define the same notion of $C^k$ iff they are compatible}. That is, all possible notions of a $C^k$ function on $M$ correspond to maximal atlases, or ``$C^k$ \textit{structures}.''

The presence of a $C^k$ structure enriches $M$ and allows one to say more about it, so it is natural to ask whether a given $M$ has a $C^k$ structure. Whitney showed that all manifolds with a $C^k$ structure also have a $C^{\infty}$ structure that can be obtained by restricting the corresponding atlas (and hence a $C^j$ structure for $j>0$). So the $C^k$ structures come together. However, there are topological manifolds with no $C^1$ structure, and hence no $C^k$ structure for any $k>0$. Thus the only distinction is between smooth manifolds and non-differentiable manifolds. We will be concerned only with the former.\\

A bijection $f:M\to N$ between smooth manifolds is called a \textit{diffeomorphism} if it and its inverse are both $C^1$. This is more strict than a \textit{homeomorphism}, which is only required to be continuous in both directions.

While in Algebraic topology we are concerned with the homotopy types of spaces, which is a coarser characterization than homeomorphism type, when studying smooth manifolds we can also ask about the diffeomorphism type, which is finer\footnote{For example, the homeomorphism type of $S^7$ splits into 28 diffeomorphism types, as shown by Milnor.}.

\subsection{Submanifolds, Immersions, Submersions, and Embeddings}
All $m$-manifolds $M$, because they are locally homeomorphic to $\R^m$, have an associated \textit{tangent space} $T_pM$ at each point $p$. When $M$ is inside a Euclidean space, this is literally the set of tangent vectors to $M$ at $p$. In general it can be defined for all manifolds $M$ in a compatible way. $T_pM$ is in fact a vector space of dimension $m$. For smooth maps $f:M\to N$, we can think of the derivative $D_pf$ as literally a linear map between tangent spaces $D_pf :T_p M \to T_{f(p)}N$.\\

The quality of $D_pf$ locally tells us a lot about its overall properties.
\begin{itemize}
	\item $f$ is called an \textit{immersion} if $D_pf$ is injective at all points $p$.
	\item $f$ is called a \textit{submersion} if $D_pf$ is surjective (i.e. full rank) at all points $p$.
	\item $f$ is called a \textit{local diffeomorphism} if $D_pf$ is invertible at all points $p$.
\end{itemize}
Note that a local diffeomorphism need not be a diffeomorphism because though it is locally invertible it might not be globally (consider e.g. the map $\R\to S^1$ given by $x\mapsto e^{ix}$).

The image of a local diffeomorphism need not even be a manifold. Consider the map taking $S^1$ to a figure-eight. Locally, every section of the circle is sent to a segment of the figure-eight, yet near the crossing point there is no homeomorphism to $\R$, so the figure-eight isn't a manifold. Another way of looking at this is that the tangent space to the figure-eight at the crossing point is \textit{not} a vector-space. For a higher-dimensional example, take the Klein bottle. There is a local diffeomorphism to a ``fake Klein bottle'' in $\R^3$ which self-intersects in a circle.

If an immersion is injective \textit{and} homeomorphic onto its image, it is called an \textit{embedding}\footnote{The homeomorphism part is necessary because for example one can construct an injective immersion $(0,1)\to \R^2$ whose image is a figure-eight.}. The image of an embedding $f:M\to N$ is always a manifold, and is called a \textit{submanifold} of $N$. Submanifolds can be equivalently characterized in another way: a manifold $A\subset N$ is a submanifold of $N$ if the charts $\kappa:N\to \R^n$ send $A$ to a linear subspace $\R^k\subset \R^n$.\\

Question: Can a given $N$ be realized as a submanifold of a given $M$ (especially when $M=\R^m$)? Or in another way, can we classify all submanifolds of a given $M$ up to diffeomorphism type?
\begin{itemize}
	\item Any smooth manifold (as long as its topology has a countable basis) can be embedded in $\R^m$ for sufficiently large $m$, though exactly what $m$ is the minimum is not always easy to determine. 
	\begin{itemize} 
		\item \textbf{Whitney's Embedding Theorem}: a smooth $m$-manifold can always be embedded in $\R^{2m}$ and immersed in $\R^{2m-1}$ (smaller dimensions may be possible). 
		\item Cohen proved more generally that a smooth $m$-manifold could be immersed in $\R^{2m-a(m)}$, where $a(m)$ is the number of 1's in the binary expansion of $m$.
		\item For example, the Klein bottle $K$ can be described without reference to an underlying space via gluing instructions, and the lowest-dimensional space it can be embedded in is $\R^4$. In $\R^3$ it can be \textit{immersed}, but not embedded.
	\end{itemize}
\item \textbf{Inverse/Implicit Function Theorem}:
\begin{itemize}
	\item If $f:M\to N$ is an immersion at $p$, $f(U)$ is a submanifold of $N$ for some open $U\ni p$.
	\item If $f:M\to N$ is a submersion at $p\mapsto q$, then the fiber $f^{-1}(q)$ is a submanifold of $M$.
\end{itemize}
\item \textbf{Transverse Intersections}:
\begin{itemize}
	\item We say $f:M\to N$ is \textit{transverse} to a submanifold $Q\subseteq N$ if for $q=f(p)\in Q$, $D_qf(T_p(M))$ and $T_q(Q)$ span $T_q(N)$. This is denoted $f\pitchfork Q$.
	\item In this case, $f^{-1}(Q)$ is a submanifold of $M$ with the same codimension as $Q$ in $N$.\\
\end{itemize}
\end{itemize}

\subsection{Tangent Bundles and Vector Fields} Every smooth manifold $M$ has an associated \textit{tangent bundle} $TM$ whose elements are pairs $(p,v)$ where $v\in T_pM$. It is a $2m$-dimensional manifold.\\

A \textit{vector field} over a manifold is a smooth map $V:M\to TM$ with $V(p)\in T_p(M)$. For some manifolds $M$, it is possible to give a basis for $TM$ by vector fields; that is, to give vector fields $V_1,V_2,\dots,V_m$ such that their values at $p\in M$ are always a basis of $T_pM$. Such manifolds are called \textit{parallelizable}.\\

Question: Which $M$ are parallelizable?
\begin{itemize}
	\item Examples:
	\begin{itemize}
		\item $M=S^1$ is parallelizable. A basis can be given by a $90$-degree rotation at every $p$. Thinking of $S^1$ as $\C^{\times}$, this corresponds algebraically to a multiplication by $i$.
		\item $M=S^2$ is not. In fact, there isn't even a nonzero vector field $V:M\to TM$. Suppose there were such $V:M\to TM$. Such $V$ induces by projection a map $V':S^2\to S^2$ with $V'(p)\bot p$ for all $p\in S^2$. But we know from Algebraic Topology that every map $S^n\to S^n$ either maps at least a point to its antipode or has a fixed point (if $p,V'(p)$ are not antipodes then there is a unique shortest path between them, so we can homotope every $V'(p)$ continuously along this shortest path to $p$, giving a homotopy to the identity map, and hence there is a fixed point).
		\item $M=S^3$ is parallelizable. It is given by analogy to the Quaternions: thinking of $S^3$ as $\HH^{\times}$, then we have a basis via the three vector fields $p\mapsto ip,p\mapsto jp,p\mapsto kp$. 
		\item As it turns out, $S^1,S^3,S^7$ are the only spheres that are parallelizable.
	
	\end{itemize}
	\item In general, all Lie groups are parallelizable (we show this on HW).\\
	\end{itemize}
	
A \textit{derivation} is an $\R$-linear map $D:\EE(U)\to \EE(U)$ (where $\EE(U)$ is the space of functions on $U$) which obeys the Leibniz rule $D(fg)=Df\cdot g  +f\cdot Dg$. The derivative along any vector is a natural example of such a map, and more generally a vector field can act on $\EE(U)$ in the same way. In fact, \textbf{all derivations correspond to actions of vector fields}; that is, every derivation is of the form
$$
V(f)(p) = (f\circ \gamma)'(0) = D_p (f)\cdot V(p) 
$$
where $\gamma:\R\to U$ is a smooth path such that $\gamma(0)=p$ and $\gamma'(p)=V(p)$ (note that this only depends on the vector field $V$).

To go the other way, the vector field associated with derivation $D$ is
$$
V(p) = (D(\gamma_1)(p),D(\gamma_2)(p),\dots,D(\gamma_m)(p))
$$
where $\gamma_j:\R\to M$ are paths with $\gamma_j(0)=p$ which form a basis of the tangent space $T_pM$.\\ 

For any derivations $V,W$ on $M$, the \textit{Lie bracket} $[V,W]:=V\circ W-W\circ V$ is a derivation (simple exercise), and hence the Lie bracket can be defined on the space of smooth vector fields on $M$, making this space into a Lie Algebra. The Lie bracket is bilinear, anti-symmetric, and satisfies the Jacobi identity:
$$
[A,[B,C]]  + [B,[C,A]] + [C,[A,B]] = 0.
$$

\subsection{Flows}
\begin{itemize}
\item A \textit{flow} is a smooth map $H:\R\times M\to M$ with the properties \begin{itemize}
	\item $H_0=\id_M$
	\item $H_tH_s=H_{t+s}$.
\end{itemize}
\item We write $H_t:M\to M$ for $H_t:p\mapsto (t,p)$. The map $t\mapsto H_t$ is a group homomorphism $\R$ to the group of diffeomorphisms on $M$. 
\item The \textit{orbit through $p$}, denoted $\gamma_p$, is $\gamma_p:t\mapsto H_t(p)$.
\item For each $H$ we have an associated vector field $V_H:p\mapsto \gamma_p'(0)\in T_pM$. This is called the \textit{infinitesimal generator} of $H$.
\begin{itemize}
	\item This correspondence is not bijective; there are vector fields that do not come from flows.
	\item However, there is a bijective correspondence between \textit{maximal local flows} and vector fields. A local flow is like a flow except it need not be defined on all of $\R\times M$, rather for each $m\in M$ it is defined on some open interval around 0. It is maximal if this domain $\Omega\subset \R\times M$ is as large as possible, so on the boundary $H$ should be diverging.
	\item The maximal local flow corresponding to $V$ is called the \textit{evolution} of $V$.
\end{itemize}
\item Question: which $V$ generate a flow (as opposed to a local flow)?
\begin{itemize}
	\item If the evolution of $V$ contains a fixed time interval $(-\ep,\ep)$ at every point of $M$, then it generates a flow (HW).
	\item If $V$ is compactly supported, then it generates a flow (HW).\\
\end{itemize}
\end{itemize}

One natural application of this concept is to describe the vector field $[V,W]$: if $V$ generates a flow $H$, then for any vector field $W$,
$$
\partial_t|_{t=0} H_t^*W = [V,W]
$$
where $H_t^*W$ is the vector field $W$ pulled back for time $t$ along $H$, i.e. $[V,W]$ is the direction and magnitude along which $W$ moves as you follow the flow of $V$.

In another way, if $V,W$ generate flows $H,I$, then we can write $[V,W]$ as a derivative in terms of these flows:
$$
\partial_s\partial_t\big|_{s=t=0} I_tH_sI_t^{-1}H_s^{-1} = [V,W].
$$
Intuitively, $[V,W]$ measures the failure of $H$ and $I$ to commute. Conversely, if $[V,W]=0$, then $H_sI_t=I_tH_s$ (HW).\\

\subsection{Vector Bundles}

A \textit{vector bundle} over $M$ is type of projection map $\pi:E\to M$ from some ``total space'' $E$ such that the fibers $E_p:=\pi^{-1}(p)$ have the structure of $r$-dimensional vector spaces. This vector space identification must be smooth in $p$, meaning precisely that $M$ can be covered by open sets $U$ such that $E_U$ is \textit{locally trivial} via $\rho:E_U\to \R^r$, which satisfies the following properties:
\begin{itemize}
	\item Restricted to a single fiber, $\rho:E_p\to \R^r$ is a vector space isomorphism.
	\item $E_U$ is diffeomorphic to $U\times \R^r$ via $(\pi,\rho):E_U\to U\times \R^r$.\\
\end{itemize}

Examples:
\begin{itemize}
	\item The trivial bundle $\R^r\times M$ with $E=\R^r$ and $\rho:\R^r\to \R^r$ the identity.
	\item The tangent bundle $TM$. 
	\item The ``tautological line bundle'' $\gamma_m^1$ over $\P^m$, where $E=\{(x,\ell):x\in \R^{m+1}\cap \ell, \ell\in \P^m\}$, and $\pi$ is the projection in the second element.\\
\end{itemize}

In each of the intersections $U_{\a}\cap U_{\b}$, at a point $p\in U_{\a}\cap U_{\b}$ there is a transition map 
$$
\rho_{\a,p}\circ \rho_{\b,p}^{-1} : \R^r\to E_{U\cap U'} \to \R^r
$$
which is an invertible linear map in $\GL_r(\R)$. Thus we have a \textit{transition function} $g_{\a,\b}:U_{\a}\cap U_{\b}\to \GL_r(\R)$ defined by $g_{\a,\b}(p) := \rho_{\a,p}\circ \rho_{\b,p}^{-1}$.\\

If we want to take a map $E\to M$ with some atlas $U_{\a},\rho_{\a}$ and make it into a vector bundle, it suffices to know that these maps $g_{\a,\b}$ are smooth. We can use this fact to define duals, tensor products, direct sums, etc. of vector bundles.\\

Some associated definitions:
\begin{itemize}
	\item A \textit{sub-bundle} of $E\to M$ is a subspace $E'\subseteq E$ for which $\rho_U(E_U\cap E')=\R^{r'}\times 0$.
	\item There is a \textit{quotient bundle} $E/E'$ between a bundle $E$ and sub-bundle $E'$.
	\item If $M\subset N$ is a submanifold, then $TM$ is a sub-bundle of $TN|_M$. The quotient bundle $\nu_{N,M}:=TN|_M/TM$ is called the \textit{normal bundle} of $M$ and $N$. 
	\item A map $s:M\to E$ such that $\pi(s(m))=m$ is called a \textit{section} of $\pi$. The space of sections is $\EE(M)$. Within a chart $\rho:E_U\to U\times \R^r$, a section can be specified by a smooth map $U\to \R^r$.\\
\end{itemize}

Using this terminology, we can view vector fields on $M$ as sections of the tangent bundle on $M$.

\newpage

\section{Differential Forms}

In this section we'll concretely define the notion of a differential (like $\d x$ or $x\d x$, etc.) using the language built up so far.\\

The \textit{cotangent space} of $M$ at $p$ is the dual of the tangent space, $T_p^*M$. Similarly, the \textit{cotangent bundle} of $M$ is the dual of the tangent bundle, $\tau_M^*:T^*M\to M$.\\

A \textit{differential} on $M$ is a section of the cotangent bundle, i.e. a map $M\to T^*M$. The basic example is $\d\ph:p\mapsto \d_p\ph$ where $\ph:M\to \R$ is some function. $\d_p\ph$ acts on a vector field $V:M\to TM$ by $\d_p(\ph)(V) = V(\ph)$. Locally (i.e. in some $U$ treated as a subset of $\R^m$),
$$
\d f = \sum_{i=1}^m \frac{\partial f}{\partial x_j} \d x_j\big|_{U}.
$$
This $\d$ is a linear map from functions on $U$ to sections of $\tau^*_M|_U$: 
$$\d:\EE(U)\to \EE(\tau^*_M|_{U}) =: \EE^1(U).$$

A \textit{differential} $k$-\textit{form} is a section of $\wedge^k\tau_M^*$. So the differential defined previously is a 1-form. We can think of a $k$-form on $M$ as a thing which can (skew-linearly) extract a real value at each point of $M$ from a $k$-tuple of $m$-dimensional vector fields.\\

The space of $k$-forms on a given chart $M\supset U\to \R^m$ is denoted $\EE^k(U)$, and we have the exterior derivative
$$
\d : \EE^k(U)\to \EE^{k+1}(U)
$$
for all $k\geq 0$. This $\d$ in general has some properties:
\begin{itemize}
	\item $\R$-linearity.
	\item \textit{Flatness}: $\d\d(\a)=0$ for all $\a$.
	\item \textit{Leibniz Rule}: $\d(\a\wedge \b) = \d\a \wedge \b + (-1)^k \a \d \b$.
	\item \textit{Pullback-Commuting}: $\d f^*(\a) = f^*(\d \a)$ for $f:M\to N$ and $\a$ a form on $N$. 
\end{itemize}

The space of all $k\leq m$ forms on $M$ at $U$, denoted $\EE^{\bullet}(U)$, is a graded-commutative $\R$-algebra with product $\wedge$.\\

Relative to a vector field $V$ on $M$, we have an ``inner contraction'' map
$$\iota_V:\EE^{k}(M)\to \EE^{k - 1}(M).$$
$\iota_V$ is defined by
$$
(\iota_V \a )(W_1,\dots,W_{K-1}) = \a(V,W_1,\dots,W_{k-1})
$$ 
where $W_1,\dots,W_{k-1}$ are other vector fields.\\

Using both $\iota_V$ and $\d$, we define the \textit{Lie Derivative} $$\LL_V = \iota_V\d + \d \iota_V : \EE^k(M)\to \EE^k(M).$$
This $\LL_V$ captures the change in a form as it flows along $V$; i.e.
$$
\frac{\d}{\d t}\Big|_{t=0} H_t^*\a = \LL_V(\a).
$$

\subsection{De Rham Cohomology} The exterior derivative yields a sequence
$$
\EE^0(M) \xra{\d} \EE^{1}(M)\xra{\d} \EE^{2}(M)\to \dots
$$
which has $\d^2=0$, so it has a defined homology and cohomology. This cohomology $$H^k_{dR}(M) := \frac{\ker(\d)}{\im(\d)}$$
is called the De Rham Cohomology. The Poincar\'e Lemma says
$$
H^k_{dR}(\R^m) = \begin{cases}
	\R& k=0\\
	0 & k>0
\end{cases}
$$
i.e. that every closed $k>0$ form on $\R^m$ is exact. And in general if $\pi:M\times \R\to M$ is the projection map, then the induced map $$H^k_{dR}(M) \xra{\pi^*} H^{k}_{dR}(M\times \R)$$
is an isomorphism. To prove this, we can show that $\d\EE^{k+1}(M\times \R)$ and $\EE^k(M)$ span $\EE^k(M\times \R)$.\\

\subsection{Stokes' Theorem} Let $\a\in \EE^m(\R^m)$ with compact support, so 
$$
\a = f(x) \d x_1 \wedge \cdots \wedge \d x_m
$$
for some compactly supported function $f:\R^m\to \R$. We define the integral of $\a$ over $\R^m$ to be 
$$
\int_{\R^m} \a := \int_{\R^m}f(x) \; \d x_1\wedge\cdots\wedge \d x_m.
$$
If we change coordinates $(x_1,\dots,x_m)\to (y_1,\dots,y_n)$ via some smooth map $h: \R^m\to \R^m$, we might integrate $h^*(\a)$, which gives
\begin{align*}
\int_{\R^m}h^*(\a) &:= \int_{\R^m} f(h(x)) \; \d h_1 \wedge \cdots \wedge \d h_m\\
&= \int_{\R^m} f(y) \cdot |\det(h)(x)|\; \d y_1\wedge \cdots \wedge \d y_m.
\end{align*}
$\det(h)$ is everywhere nonzero because it is smooth. If in addition it is positive everywhere (which we can assume as long as the support of $f$ is connected), then 
$$
\int h^*(\a) = \int \a.
$$

\end{document}