\documentclass{amsart}

\input{C:/Users/jzchr/MathDocuments/preamble_general.tex}

\title{Math 317 HW 7}
\author{Jalen Chrysos}

\begin{document}
	\maketitle
	\textbf{Problem 1 (Hatcher 4.1.5)}: For a pair $(X,A)$ of path-connected spaces, show that $\pi_1(X,A,x_0)$ can be identified in a natural way with the set of cosets $\a H$ of the subgroup $H\subset \pi_1(X,x_0)$ represented by loops in $A$ at $x_0$.
	\begin{proof}
		By definition, $\pi_1(X,A,x_0)$ is the group of (homotopy classes of) maps $\gamma: I\to X$ beginning at $x_0$ and ending in $A$, where the homotopies are rel $x_0,A$. Since $A$ is connected, such $\gamma$ can be homotopically extended to end at $x_0$ and thus form a loop. All possible extensions within $A$ are homotopic to $\gamma$, and thus are all equivalent in $\pi_1(X,A,x_0)$.\\
		
		The correspondence is as follows: each class $[\gamma] \in \pi_1(X,A,x_0)$ contains loops in $\pi_1(X)$. If $\a$ is such a loop in $[\gamma]$, then all of $\a H$ is in $[\gamma]$ because concatenating loops in $A$ (elements of $H$) can be done homotopically rel $A$. So this $\a H$ is the class we associate to $[\gamma]$. It is well-defined because if $\b\simeq \a$ in $\pi_1(X)$, then the homotopy between them fixes both endpoints at $x_0$ which is strictly stronger than fixing one and keeping the other in $A$, so $\b\simeq \a$ in $\pi_1(X,A,x_0)$ as well. 
		
		The inverse is just given by inclusion of $\a H$ into $\pi_1(X,A,x_0)$, as they are all in the same class (as shown above). It is easy to check that the composition of these two is the identity.
		
%		In each of those classes there is a loop at $x_0$, since $A$ is path connected and paths can be extended by homotopy.
%		
%		If $\a$ is a loop at $x_0$ in $X$, then everything in $\a H$ is homotopic to $\a$ (do later)
	\end{proof}
	
	\newpage
	
	\textbf{Problem 2 (Hatcher 4.1.8)}: Show that the sequence 
	$$
	\pi_1(X,x_0) \xrightarrow{f} \pi_1(X,A,x_0) \xrightarrow{\partial} \pi_0(A,x_0) \xrightarrow{g} \pi_0(X,x_0)
	$$
	is exact.
	\begin{proof}
		$\ker(\partial)$ is exactly the set of paths in $\pi_1(X,A,x_0)$ which are loops, equivalently the paths whose endpoint is $x_0$, equivalently the image of $\pi_1(X,x_0)$.\\
		
		Next, $\im(\partial)$ is the points in $A$ that are the other endpoint of a path in $X$ from $x_0$, i.e. the points in the same connected component of $X$ as $x_0$. And this is also the kernel of $g$, so indeed the sequence is exact here too.
	\end{proof}
	
	\newpage 
	\textbf{Problem 3 (Hatcher 4.1.15)}: Show that every map $f:S^n\to S^n$ is homotopic to a multiple of the identity map by the following steps:
	\begin{enumerate}[(a)]
		\item Use Lemma 4.10 to reduce to the case that there is a point $q\in S^n$ with $f^{-1}(q)=\{p_1,\dots,p_k\}$ and $f$ is an invertible linear map near each $p_i$.
		\item For $f$ as in (a), consider the composition $gf$ where $g:S^n\to S^n$ collapses the complement of a small ball about $q$ to the basepoint. Use this to reduce (a) further to the case $k=1$.
		\item Finish the argument by showing that an invertible $n\times n$ matrix can be joined by a path of such matrices to either the identity matrix or the matrix of a reflection. 
	\end{enumerate}
	\begin{proof}
		(a): Applying lemma 4.10 in the case $Z=S^n$ and $W=\emptyset$ says that $f$ is homotopic to a map $f_1:S^n\to S^n$ which is piecewise linear (viewing $S^n$ as $\R^n$ locally) on some polyhedron $K\subset S^n$ such that $K$ contains $f_1^{-1}(U)$ for some $U\subset \R^n$. Therefore, it suffices to show the case where $f$ is of this form.
		
		Since $f$ is piecewise linear on $K$ (with finitely many pieces), the points $p$ defining the corners of linear pieces can have $f(p)$ moved slightly while keeping $f$ linear on the adjoining pieces, and this is clearly continuous. We can homotope $f$ by adjusting its value on these points $p$ so that all the piecewise linear parts become invertible. Then any given point $q\in U$ can have at most one preimage per linear piece and none outside of $K$, so the preimage is finite. And one can slightly move $q$ such that the preimages are all strictly inside the pieces, achieving the desired conditions.\\
		
		(b): Let $|f^{-1}(q)|=k$. Outside $k$ small neighborhoods of preimages of $q$, the image of $f$ is contained in $S^n\setminus \{q\}\simeq D^n$, which is contractible, so we can homotope $f$ so that it is the basepoint outside of these $k$ neighborhoods. But then $f$ is literally a product (in the $\pi_n$ sense) of $k$ maps, each of which has a single preimage of $q$. Thus if we show that any $f'$ with $k=1$ is of the form $\pm \id \in \pi_n$, then this $f$ will be $\pm k\cdot \id$, which proves the theorem for general $k$. So it suffices to address the case $k=1$.\\
		
		(c): Suppose we have a map $f$ of this type with $k=1$. In the same way as part (b), we can homotope $f$ so that it is constant outside a neighborhood of $f^{-1}(q)$, and we already know it is invertible inside such a neighborhood from part (a). If we can deform $f$ to $\pm\id$ in this neighborhood, then that will prove the theorem.
		
		The space of $n\times n$ matrices with \textit{positive} determinant is path-connected, as a subspace of $\R^{n^2}$. One way to see this is that every matrix $a\in M_n(\R)$ has a well defined logarithm
		$$
		\log(1 + a) = a + a^2/2 + a^3/3 + a^4/4 + \cdots
		$$
		and exponential 
		$$
		e^{a} = I + a + a^2/2 + a^3/6 + a^4/24 + \cdots
		$$
		so one can let $x=\log(a)$ and take the path $\gamma(t) = e^{tx}$. Then $\gamma(0)=I$, $\gamma(1)=e^x=a$, and $e^{tx}\in \GL_n(\R)$ because it has inverse $e^{-tx}$. 
		
		Similarly, the space of negative determinant matrices is path connected. Whichever one of these we are in initially, we can deform to get $\id$ or $-\id$. 
	\end{proof}
	
	\newpage
	\textbf{Problem 4 (Hatcher 4.1.17)}: Show that if $X$ and $Y$ are CW complexes with $X$ $m$-connected and $Y$ $n$-connected, then $(X\times Y, X\vee Y)$ is $(m+n+1)$-connected, as is the smash product $X\wedge Y$.
	\begin{proof}
		We have the long exact sequence
		$$
		\cdots \to \pi_j(X\vee Y) \to \pi_j(X\times Y)\to \pi_j(X\times Y,X\vee Y) \to \pi_{j-1}(X\vee Y)\to \cdots 
		$$
		
		$X$ is homotopy equivalent to a CW complex $X'$ with one 0-cell and no other cells of dimension $\leq m$, and likewise $Y$ has a corresponding $Y'$. $(X\times Y,X\vee Y)\simeq (X'\times Y',X'\vee Y')$ so it suffices to show this for the latter. 
		
		The cells of $X'\times Y'$ are the same as $X'\vee Y'$ in small dimensions. The smallest-dimension cells in $X'\times Y'$ not in $X'\vee Y'$ are of dimension $m+n$, so $\pi_j(X'\vee Y')\to \pi_j(X'\times Y')$ is automatically an isomorphism for $j\leq m+n$, and thus in the long exact sequence
		$$
		\cdots \to \underbrace{\pi_{m+n+1}(X' \times Y',X'\vee Y')}_{0} \to \pi_{m+n}(X'\vee Y') \cong \pi_{m+n}(X'\times Y') \to \underbrace{\pi_{m+n}(X' \times Y',X'\vee Y')}_{0}\to \cdots
		$$
		implies $\pi_{m+n+1}(X' \times Y',X'\vee Y')=0$ and likewise for all smaller dimensions.\\
		
		The smash product $X\wedge Y$ is the quotient $X\times Y/X\vee Y$. Excision for homotopy groups implies that 
		$$
		\pi_j(X\times Y / X\vee Y) \cong \pi_j(X\times Y, X\vee Y)
		$$
		for $j\leq m+n$. For $j=m+n+1$, the inclusion induces a surjection, and the only thing 0 can surject onto is 0, so $\pi_{m+n+1}(X\wedge Y)=0$ as well.
	\end{proof}
	
	\newpage
	\textbf{Problem 5 (Hatcher 4.2.1)}: Use homotopy groups to show that there is no retraction $\RP^n\to \RP^k$ for $n>k>0$.
	\begin{proof}
		$\RP^n$ has a 2-sheeted cover from $S^n$, as we've seen before. So $\pi_k(\RP^n)=\pi_k(S^n)=0$, but $\pi_k(\RP^k)=\pi_k(S^k)=\Z$ (assuming $k\geq 2$). And in the case $k=1$, $\RP^1=S^1$, so it is still true that $\pi_1(\RP^1)=\Z$. A retraction would induce an isomorphism between $\pi_k(S^n)$ and $\pi_k(S^k)$, but they are not isomorphic, so there is no retraction.
	\end{proof}
	
	\newpage
	\textbf{Problem 6 (Hatcher 4.2.5)}: Let $f:S^2_{\a}\vee S^2_{\b} \to S^2_{\a}\vee S^2_{\b}$ be the map which is the identity on $S^2_{\a}$, and the sum of the identity and a map $S^2_{\b}\to S^2_{\a}$ on $S^2_{\b}$. Let $X$ be the mapping torus of $f$, i.e.
	$$
	X := \frac{(S^2_{\a} \vee S^2_{\b} )\times I}{(x,0)\sim (f(x),1)}
	$$	
	The mapping torus of the restriction of $f$ to $S^2_{\a}$ forms a subspace $A=S^1\times S^2_{\a}\subset X$. Show that the maps $\pi_2(A)\to \pi_2(X)\to \pi_2(X,A)$ form a short exact sequence 
	$$
	0 \to \underbrace{\pi_2(A)}_{\Z}\to \underbrace{\pi_2(X)}_{\Z^2}\to \underbrace{\pi_2(X,A)}_{\Z}\to 0
	$$
	and compute the action of $\pi_1(A)$ on these three groups. In particular show that the action is trivial on $\pi_2(A)$ and $\pi_2(X,A)$ but nontrivial on $\pi_2(X)$.
	\begin{proof}
		First, we already know that the sequence
		$$
		\pi_3(X,A)\to \pi_2(A)\to\pi_2(X)\to\pi_2(X,A)\to\pi_1(A)\to\pi_1(X)\to\pi_1(X,A)
		$$
		is exact. To show the short exactness of this piece, it suffices to show that $\pi_1(A)\to \pi_1(X)$ is an isomorphism, which is clear, and that $\pi_3(X,A)=0$. Note that this incidentally shows $\pi_1(X,A)=0$, which we will use later.
		
		Now let's calculate each of these groups $\pi_2$. We can immediately see that 
		$$
		\pi_2(A) = \pi_2(S^1\times S^2) = \pi_2(S^1)\times \pi_2(S^2) = \Z.
		$$
		Similarly $\pi_2(X,A)$ must be $\pi_2(X/A)=\Z$ as well by Hurewicz, because $\pi_1(X,A)=0$ (we will show next that the action of $\pi_1(A)$ is trivial). And it follows by exactness of the sequence above that $\pi_2(X)$ must be $\Z^2$.
		
		To compute the action of $\pi_1(A)$, consider the universal cover $p:X'\to X$. $X'$ is the mapping torus rolled out into a line. $\pi_1(A)$ acts on this space by translating the line. 
		
		For $\pi_2(A)$, the action is trivial because the 2-sphere is just being pushed down the $S^2_{\a}\times \R$ tube, and can be retracted back again homotopically. Similarly for $\pi_2(X,A)$ with the $S^2_{\b}\times \R$ tube. But with $\pi_2(X)$, the action of $\pi_1(A)$ can split a sphere in $S^2_{\b}$ between the two tubes, so it is nontrivial.
	\end{proof}
	\newpage
	\textbf{Problem 7 (Hatcher 4.2.7)}: Construct a CW complex with prescribed homotopy groups and prescribed actions of $\pi_1$ on those homotopy groups.
	\begin{proof}
		One can construct, by taking a product of $K(G_n,n)$ for $n\geq 2$, a space $X'$ which has homotopy groups $\pi_n(X)=G_n$ for $n\geq 2$. Now we want $X'$ to be a covering space of some $X$, $p:X'\to X$ (which necessarily has the same homotopy groups for $n\geq 2$) while making $\pi_1(X)=G_1$ \textit{and} ensuring the desired actions of $G_1$ on each $G_n$. It suffices to do this for one $n$ at a time.\\
		
		We just need the deck group of this covering to match the desired action of $\pi_1$. We can attach a contractible tree of 1-cells representing the free group, and then in $X$ we can identify some of the branches where there are relations that want to impose.
	\end{proof}
	
	\newpage
	\textbf{Problem 8 (Hatcher 4.2.10)}: Let $X$ be the CW complex obtained from $S^1\vee S^n$ (where $n\geq 2$) by attaching $e^{n+1}$ by a map representing the polynomial $p(t)\in \Z[t,t^{-1}]\cong \pi_n(S^1\vee S^n)$, so that $\pi_n(X)\cong \Z[t,t^{-1}]/(p(t))$. Show that $\pi_n'(X)$ is cyclic and compute its order in terms of $p(t)$. Give examples showing that the group $\pi_n(X)$ can be finitely-generated or not, independently of whether $\pi_n'(X)$ is finite or infinite.
	\begin{proof}
		$S^1\vee S^n$ has a universal cover $Y$ that is a line with $\Z$ copies of $S^n$ attached along it. $\pi_1$ acts on $Y$ by shifting this line.\\
		
		$\pi_n'(X)$ is $\pi_n(X)$ but quotienting out the action of $\pi_1(X)$. Since $\pi_1(X)$ acts as multiplication by $t$, this is the same as setting $t$ to 1, or quotienting by $(t-1)$. That is,
		$$
		\pi_n'(X) = \Z[t,t^{-1}]/(p(t),t-1) = \Z/(p(1))
		$$
		So $\pi_n'(X)$ is cyclic of order $p(1)$ (or infinite if $p(1)=0$).\\
		
		$\pi_n'(X)$ finite, $\pi_n(X)$ finitely-generated: take $p(t)=1-t-t^2$, for which $\pi_n(X)$ can be generated by just $1$ and $t$, and $\pi_n'(X)=0$.\\
		
		$\pi_n'(X)$ finite, $\pi_n(X)$ infinitely-generated: take $p(t)=2-t$, for which $\pi_n(X)=\Z[t,t^{-1}]/(2-t) = \Z[\tfrac12]$ which is infinitely generated, but $\pi_n'(X)=0$.\\
		
		$\pi_n'(X)=\Z$, $\pi_n(X)$ finitely-generated: take $p(t)=t-1$, for which $\pi_n(X)=\pi_n'(X)=\Z$.\\
		
		$\pi_n'(X)=\Z$, $\pi_n(X)$ infinitely-generated: take $p(t)=2t-2$, for which $\pi_n'(X)=\Z$ and $\pi_n(X)$ has nonzero and unrelated $t^n$ for all $n\in \Z$, so it is infinitely-generated. 
	\end{proof}
	
	\newpage
	\textbf{Problem 9 (Hatcher 4.2.13)}: Show that a map between connected $n$-dimensional CW complexes is a homotopy equivalence if it induces an isomorphism on $\pi_i$ for $i\leq n$.
	\begin{proof}
		If $f_*:\pi_i(X)\to\pi_i(Y)$ is an isomorphism for $i\leq n$ then $\pi_i(Y,X)=0$ for $1<i\leq n$. The case $i=1$ may not be true. But if we pass to universal covers $Y'$ and $X'$, then $\pi_i(Y',X')=0$ for $1\leq i \leq n$, so $H_i(Y',X')=0$ by Hurewicz Theorem. And for $i> n$, $H_i(Y',X')=H_i(Y'/X')=0$ automatically because $Y'/X'$ has no $i$-cells, as $Y',X'$ are $n$-dimensional. Thus, by Hurewicz again, $\pi_i(Y',X')=\pi_i(Y,X)=0$ for all $i$, and so $f_*$ induces an isomorphism in all dimensions, and by Whitehead's theorem $f$ is therefore a homotopy equivalence. 
	\end{proof}
	
	
\end{document}