\documentclass{amsart}
\input{C:/Users/jzchr/MathDocuments/preamble_general.tex}

\title{MATH 318 HW 1}
\author{Jalen Chrysos}

\begin{document}
	
	\maketitle
% OPTIONAL QUESTIONS
%\textbf{Problem 1 (2.3)}: Prove that two $C^k$-manifolds (with $k\geq 1$) that are $C^k$-diffeomorphic have the same dimension.
%
%\newpage 
%
%\textbf{Problem 2 (2.5)}: Prove that for $m\leq n$, $S^m$ is a smooth submanifold of $S^n$ if we identify $S^m$ with $S^n\cap (\R^{m+1}\times 0)$. Conclude the same for $P^m$ and $P^n$.
%
%\newpage 

\textbf{Problem 1 (2.6)}: Let $f:\R^n\to \R$ be a homogeneous function of degree $d$. Prove that $f^{-1}(1)$ is a (possibly empty) submanifold of dimension $n-1$.
\begin{proof}
	By Example 2.6, it suffices to show that the derivative of a homogeneous function $f:\R^n\to \R$ is surjective on points within $f^{-1}(1)$. In this case because the dimension of the output space is 1, it is equivalent to show that the derivative is nonzero. 
	
	For $f$ to be homogeneous of degree $d$ means that $f(\lambda v)=\lambda^df(v)$ for all $v\in \R^n$ and $\lambda \in \R$. In particular, where $f(v)=1$, $f(\lambda v) = \lambda^df(v) = \lambda^d$, so the derivative in direction $v$ (which is defined because $f(0)=0$ so $0\not\in f^{-1}(1)$) at $v$ is nonzero. In particular 
	$$\partial_{t} f(v+t v) = \partial_{t} f((1+t)v) = \partial_{t} (1+t)^d = d(1+t)^{d-1} = d \; \text{(at $t=0$)}$$ 
	so $f$ has nonzero derivative at all points in $f^{-1}(1)$, thus making $f^{-1}(1)$ a submanifold of dimension $n-1$ as desired.
\end{proof}

\newpage

\textbf{Problem 2 (2.7)}: Show that $\SL_n(\R)$ is a smooth submanifold of $\R^{n^2}$ and determine its dimension. Prove also that the map $\SL_n(\R)\times \SL_n(\R)\to \SL_n(\R)$ via $(\sigma,\tau)\mapsto \sigma\tau^{-1}$ is smooth. Do the same for $\SO_n(\R)$.
\begin{proof}
	$\SL_n(\R)$ is the preimage $\det^{-1}(1)$ of $\det:\R^{n^2}\to \R$. The determinant is homogeneous of degree $n$. Thus, by the previous problem, $\SL_n(\R)$ is a submanifold of dimension $n^2-1$.
	
	$\tau\mapsto \tau^{-1}$ is smooth on $\SL_n(\R)$ as it is given by the adjugate matrix, so each coordinate is just the determinant of one of the minors, a polynomial in the matrix entries in $\tau$ and thus smooth. Similarly, each entry of $\sigma\tau^{-1}$ is a polynomial in the entries of $\sigma$ and $\tau$ and thus smooth.\\
	
	$\SO_n(\R)$ can be seen as the fiber $f^{-1}(I)$ where $f:\GL_n(\R)\to \Sym_n(\R)$ is defined $f:a \mapsto a a^{\top}$, with the codomain being the symmetric matrices. So by Corollary 4.3, it suffices to show that $D_pf$ is surjective at every special orthogonal matrix $p$.
	
	To do this, wlog suppose we want to show that the symmetric matrix with 1 in the $(i,j)$ and $(j,i)$ places and 0 elsewhere is in the image of $D_pf$ (the diagonal case is easy). Let $c_k$ be the $k$th column of $p$. Assume wlog that $c_j$ is not in the direction $(1,1,\dots,1)$; otherwise swap $i,j$ and the following construction will work (note that they cannot both be in the same direction because $p$ has determinant 1). Consider $D_p(f)$ of the vector $v$ that perturbs the matrix by 
	$$p_{i1} \mapsto p_{i1} + v_1, \;\; p_{i2}\mapsto p_{i2} + v_2, \;\; \dots \;\; p_{in}\mapsto p_{in} + v_n.$$
	 We can choose $v$ to be orthogonal to $c_k$ for $k\neq i,j$ and also orthogonal to $(1,1,\dots,1)$. Due to this choice, the perturbation by $v$ doesn't affect $c_i\cdot c_k$ for any $k\neq i,j$, nor any dot product that doesn't involve $c_i$. $c_i\cdot c_i$ changes by $2(v_1+v_2+\cdots+v_n)$, which is 0 because $v$ to be orthogonal to $(1,1,\dots,1)$. Now, because we assumed that $c_j$ is not in the direction $(1,1,\dots,1)$, and also $p$ is orthogonal so it is not in the direction of any other $c_k$, $c_i\cdot c_j$ is changed by $v\cdot c_j$, which is nonzero. Thus, this (and its symmetric opposite) is the only entry that has nonzero determinant. So $D_pf$ is surjective, as desired.
	
	Proving that the product and inverse are smooth is the same as for $\SL_n(\R)$.
	
	As for the dimension, 4.3 gives that the codimension should be $\dim(\Sym_n(\R))$, which is $n + (n-1) + \cdots + 1 = \tfrac12(n)(n+1)$, so $\dim(\SO_n(\R)) = n^2 - \tfrac12(n)(n+1) = \tfrac12(n)(n-1)$. 
	
%	$\SO_n(\R)$ is homeomorphic to
%	$$S^{n-1}\times S^{n-2}\times \cdots \times S^1$$
%	as each $(n-1)$-tuple $(a_1,a_2,\dots,a_{n-1})$ in this product can be uniquely and continuously mapped to a $\SO_n(\R)$ matrix as follows: take the first column to be $a_1$, let the $(n-1)$-plane orthogonal to $a_1$ be $V_1$, take the second column to be $a_2\in V_1$, let the $(n-2)$-plane orthogonal to both $a_1$ and $a_2$ be $V_2$, $\dots$, and finally when $V_{n-1}$ is a 1-dimensional space, take whichever of the two unit vectors will make the resulting matrix det 1 (the other will give det -1). 
%	
%	This is continuous and bijective. This parametrization gives the dimension of $\SO_n(\R)$ as
%	$$
%	\dim(\SO_n(\R)) = (n-1) + (n-2) + \cdots + 1 = \tfrac12 n(n-1).
%	$$
\end{proof}

\newpage 

\textbf{Problem 3 (4.2)}: Find an embedding of $S^n\times S^m$ in $\R^{n+m+1}$.
\begin{proof}
	Since $S^n$ and $S^m$ are compact, it suffices to produce an injective immersion. Take $S^n$ and $S^m$ to be the submanifolds of $\R^{n+1},\R^{m+1}$ given by
	$$
	S^n = \{x_1,\dots,x_{n+1} \; : \; x_1^2 + \cdots + x_{n+1}^2 = 1\}, \;\;\; 	S^m = \{z_1,\dots,z_{m+1} \; : \; z_1^2 + \cdots + z_{n+1}^2 = 1\}. 
	$$
	Fixing some $R>1$, define the map $f:\R^{n+1}\times \R^{m+1} \to \R^{n+m+1}$ by
	$$
	f:(x_1,\dots,x_{n+1},z_1,\dots,z_{m+1}) \mapsto \Big(x_1(R+z_1),\dots,x_{n+1}(R+z_1),z_2,z_3,\dots,z_{m+1}\Big).
	$$
	$f$ is polynomial in every coordinate and thus smooth. I claim that $f$ is injective as a map restricted to $S^n\times S^m$.
	
	Suppose $(x_1',\dots,x_{n+1}',z_1',\dots,z_{m+1}')$ is another point in $S^n\times S^m$ mapped to the same output by $f$. The first $n+1$ coordinates give
	$$
	x_j(R+z_1) = x_{j}'(R+z_1') \implies \frac{x_j}{x_j'} = \frac{R+z_1'}{R+z_1} =: \lambda
	$$
	for all $1\leq j\leq n+1$. Because $R>1$ and $|z_1|,|z_1'|\leq 1$, we have $\lambda > 0$. But then
	$$
	1= x_1^2 + \cdots + x_{n+1}^2 = \lambda^2(x_1'^2 + \cdots + x_{n+1}'^2) = \lambda^2 \implies \lambda = 1. 
	$$
	This gives $z_1=z_1'$ and hence $x_j=x_j'$ for all $j$. From the remaining $m$ coordinates, it immediately follows that $z_j=z_j'$ for $j\geq 2$, so the two points are indeed equal. That is, $f$ is injective on $S^n\times S^m$, and thus it is an embedding of $S^n\times S^m$ into $\R^{n+m+1}$.
	
	%Then immediately $z_j'=z_j$ for $j\geq 2$, and hence because $z_1^2 = 1 - z_2^2-\cdots-z_{m+1}^2$ we have $z_1^2=z_1'^2$, giving $z_1=\pm z_1'$.
\end{proof}
	
\end{document}